Light inextensible strings under tension in the Schwarzschild geometry

### 2019-08-28 string_physics_teacher.tex
work

Dynamic definitions of straight line also run into difficulties:
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\cite{abramowicz1992,abramowicz1993}.

The Schwarzschild metric

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right),
\end{equation}

\noindent where units in which~$G=c=1$ are used, has been known for
over a century~\cite{schwarzschild1916}.  Here I consider the spatial
component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
throughout.  A circular string at constant~$r$ will, trivially, have
proper length~$2\pi r$.

@@

This would indicate that, close to the Schwarzschild radius $r=2M$,
large (but finite) amounts of string are required to connect two
points with slightly differing radial coordinates.  Passing to the
case $w\gg 1$ we have an asymptotic expansion that begins

@@

\noindent suggesting that there is ``extra radial space'', of length
over $1.19\cdot 2M$, sensible at large distances from the event
horizon.  Radial strings have proper length longer than the
Schwarzschild radial coordinate difference $w$, a nice illustration of
the failure of Euclidean geometry.

@@

\noindent where dashes denote differentiation with respect to~$\phi$.  The
Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give, after simplification, an expression for the
second derivative of~$r$:

% \begin{equation}\label{rdashdash_properunits}
%   r''\left(\phi\right) =
%   (r-1) + \frac{(4r-3)\left(r'\right)^2}{2r\left(r-1\right)},\qquad r>1.
% \end{equation}
\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2M) + \frac{(4r-6M)\left(r'\right)^2}{2r\left(r-2M\right)},\qquad r>2M.
\end{equation}

@@

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
circular orbits cannot follow a taut string at closest approach (as in
the classical case).  Also observe that null geodesics have a radius
of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
follow a taut string at closest approach either.  The string has a
point of inflection if~$\left(r'\right)^2=2r(r-2M)$.

@@

Here I solve equation~\ref{rdashdash} using the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  Non-radial
solutions fall into two distinct classes: those with $r>2M$ at all
points, and those which attain $r=2M$, in which case the string is
tangential to the event horizon.

We first consider strings for which $r>2M$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.
As discussed above, if $\left(r'\right)^2=2r(r-2M)$, the string has a
point of inflection at which point the radius of curvature becomes
infinite.  Figure~\ref{curvature_switch} shows non-self intersecting
strings rotated so that the point of inflection occurs at~$\phi=0$; we
see that each string has a finite length in which it is convex to the
origin and two semi-infinite stretches in which the curvature is
negative.

@@

Light inextensible strings under tension adopt non-trivial
configurations including self-intersecting paths.  The path of such
strings interprets the concept ``straight line'' in general
relativity, others being null geodesics and world lines of massive
objects.  Comparing a world line with that of nearby taut strings thus
furnishes another insight into path curvature, complementing the
analysis of Abramowicz~\cite{abramowicz1992}.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geqslant< 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

>>>>

\section{The Schwarzschild metric}

The vacuum Einstein field equations for a static, spherically
symmetric mass are solved by the Schwarzschild
metric~\cite{schwarzschild1916}.  In units in which~$G=c=1$, this is:

\begin{equation}\label{schwarzschild}
ds^2= -dt^2\left(1-2M/r\right) +\frac{dr^2}{1-2M/r} + r^2\left(d\theta^2 + \sin^2\theta d\phi^2\right).
\end{equation}

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
circular string at constant~$r$ will, trivially, have proper
length~$2\pi r$.

@@

(compare the flat space string length of $2Mw$).  This would indicate
 that, close to the Schwarzschild radius $r=2M$, large (but finite)
 amounts of string are required to connect two points with slightly
 differing radial coordinates.  Passing to the case $w\gg 1$ we have
 an asymptotic expansion that begins

@@

\noindent suggesting that there is ``extra space'', sensible at large
distances from the event horizon, with radial strings being
arbitrarily longer than flat space would allow: a nice illustration of
the failure of Euclidean geometry.

@@

\noindent where dashes denote differentiation with respect to~$\phi$.  The
Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give, after simplification\footnote{The Mathematica
  idiom would be:

\noindent
{\tt
Needs["VariationalMethods`"]\\
p = EulerEquations[Sqrt[y'[x]\^{}2/(1-1/y[x]) + y[x]\^{}2],y[x],x] // FullSimplify\\
ss = p[[1]]//Numerator\\
DSolve[ss==0,y[x],x]
}}, an expression for the second derivative of~$r$:

\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2M) + \frac{(4r-6M)\left(r'\right)^2}{2r\left(r-2M\right)},\qquad r>2M.
\end{equation}

@@

\noindent which would give~$R$ as a function of~$\left(r,r'\right)$.
Considering~$r'(\phi)=0$ shows that if a string is tangential at a
point with~$r_0>2M$, then it has a radius of curvature of~$r_0^2/2M$
at that point.  Any circular path, including a circular orbit
at~$r=r_0$, has a radius of curvature of $r_0<r_0^2/2M$, showing that
circular orbits cannot follow a taut string at closest approach (as in
the classical case).  Also observe that null geodesics have a radius
of curvature of $r_0^2/3M < r_0^2/2M$, showing that light cannot
follow a taut string at closest approach either: tangential string
looks curved to a local observer.

@@

Here I solve equation~\ref{rdashdash} numerically, using the
\verb+deSolve+ package for ordinary differential
equations~\cite{soetart2010}.  Non-radial solutions fall into two
distinct classes: those with $r>2M$ at all points, and those which
attain $r=2M$, in which case the string is tangential to the event
horizon.

We first consider strings for which $r>2M$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

Equation~\ref{roc} shows that the string has a point of inflection
if~$\left(r'\right)^2=2r(r-2M)$ at which point the radius of curvature
becomes infinite.  Figure~\ref{curvature_switch} shows non-self
intersecting strings rotated so that the point of inflection occurs
at~$\phi=0$; we see that each string has a finite length in which it
is convex to the origin and two semi-infinite stretches in which the
curvature is negative.

@@

Light inextensible strings under tension adopt non-trivial
configurations including self-intersecting paths.  The path of such
strings interprets the concept ``straight line'' in general
relativity, others being null geodesics and world lines of massive
objects.  Comparing light paths with that of nearby taut strings thus
furnishes another insight into path curvature, complementing the
analysis of Abramowicz~\cite{abramowicz1992}.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=170mm]{closest_approach_selfintersecting.pdf}
\caption{Self-intersecting strings shown for $\phi\geqslant 0$.  Closest
  approach to the event horizon occurs at $\phi=0$ (3 o'clock) and
  $r/2M\in\left(1,1.0761\right)$; strings are symmetrical about $\phi=0$
  but for clarity only $\phi\geqslant 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}
